Algorithms for Estimating Trends in Global Temperature Volatility

### 2018-11-14 aaai-2018/khodadadi-mcdonald-aaai19-camera.tex
final proofing of camera-ready

%Arash
\usepackage{amsmath,amsthm}
\usepackage{amssymb}
\usepackage{mathrsfs}
\DeclareMathOperator*{\argmin}{argmin}
\usepackage{siunitx}
\usepackage{comment}
\usepackage{mathtools}
\mathtoolsset{showonlyrefs}

@@

@@

Trends in terrestrial temperature variability are perhaps more relevant for species
viability than trends in mean
temperature~\cite{huntingford_no_2013},
because an increase in
temperature variability will increase the probability of extreme hot
outliers~\cite{VasseurDeLong2014}. Recent climate literature
suggests that it is more difficult for society to adapt to these
extremes than to the gradual increase in the mean temperature
\cite{hansen_perception_2012,huntingford_no_2013}. Furthermore, the willingness of popular media to
emphasize the prevalence extreme cold events coupled with a
fundamental misunderstanding of the relationship between climate (the global
distribution of weather over the long run) and weather (observed
short-term, localized behavior) leads to public misunderstanding of climate
change. In fact, a point of active debate is the extent to which the observed
increased frequency of extreme cold events in the 
northern hemisphere can be attributed to increases in temperature
variance  rather than to changes in
mean climate~\cite{Screen2014,FischerBeyerle2013,TrenberthZhang2014}.

Nevertheless, research examining trends in the volatility of
spatio-temporal climate data is scarce. \citet{hansen_perception_2012} studied the change in
the standard deviation (SD) of the surface temperature in the NASA
Goddard Institute for Space Studies gridded temperature dataset by
examining the empirical SD at each spatial location relative to that
location's SD over a base period and showed
that these estimates are increasing.
\citet{huntingford_no_2013} took a similar
approach in analyzing the ERA-40 data set. They argued that, while there
is an increase in the SDs from 1958-1970 to 1991-2001, it
is much smaller than found in
\cite{hansen_perception_2012}. \citet{huntingford_no_2013} also computed the
time-evolving global SD from the detrended time-series at each
position and argued that the global SD has been
stable.

@@

While the motivation for our methodology is its application to large, gridded climate
data, we note that our algorithms are easily generalizable to
spatio-temporal data under convex loss, e.g. exponential family
likelihood. Furthermore the spatial structure can be broadly construed
to include general graph dependencies. Our current application uses
Gamma likelihood which lends itself well to modeling trends in
pollutant emissions or in astronomical phenomena like microwave
background radiation. Volatility estimation in natural gas/oil markets
or with financial data is another possibility. These methods can also
be applied to resting-state fMRI data (though the penalty structure changes).

@@

Inspired by the $\ell_1$-trend filtering algorithm, we propose a
non-parametric model for estimating the variance of a time-series. To
this end, we assume that at each time step $t$, there is a parameter
$h_t$ such that the observations $y_t$
are independent normal variables with zero mean and variance
$\exp(h_t)$. The negative log-likelihood of the observed data in this
model is $l(y\given h) \propto -\sum_t h_t - y_t^2e^{-h_t}$. Crucially,
we assume that the parameters $h_t$ vary smoothly and estimate them by
minimizing the penalized, negative
log-likelihood:  
\begin{equation}
\min_h -l(y\given h)+\lambda \norm{ D_th }_1
\label{eq:l1tf_var}
\end{equation}
where $D_t$ has the same structure as above.

@@

\begin{comment}
\begin{algorithm}[tb]
  \caption{PDIP for $\ell_1$ variance estimation}
  \label{alg:pdip}
  \begin{algorithmic}
    \REQUIRE $\lambda>0$, $w>0$, $\nu\leftarrow 0$, $\mu_1\leftarrow
    0$, $\mu_2\leftarrow 0$, $J\in\mathbb{Z}^+$, $\{w_k\}$
    \COMMENT{Initialization} 
    \FOR[Central path]{$k=1,2,\ldots$}
    \FOR[Newton updates]{$j=1,\ldots,J$}
    \STATE Solve $A [\Delta\nu\; \Delta\mu_1\;
    \Delta\mu_2]^\top=r_{w_k}$ to find the search direction. 
    \STATE $\quad\quad A$ is given in \ref{eq:delta_r}
    \STATE Update $[\nu^{j+1}\; \mu^{j+1}_1\; \mu^{j+1}_2] \leftarrow
    [\nu^{j}\; \mu^{j}_1\; \mu^{j}_2] + [\Delta\nu\; \Delta\mu_1\;
    \Delta\mu_2]$. 
    \ENDFOR
   \ENDFOR
   \RETURN $h=\log\frac{y^2}{1+D^\top\nu}$ 
   \\
   \attn{Is this right? Explicit form of $A$}\\
   \attn{Arash: see appendix A}
  \end{algorithmic}
\end{algorithm}
\end{comment}

@@

At a specific time $t$, the data are measured on a grid of points with
$n_r$ rows and $n_c$ columns for a total of $S=n_r\times n_c$ spatial
locations. Let $y_{ijt}$ denote the value of the 
observation at time $t$ on the $i^\text{th}$ row and $j^\text{th}$
column of the grid, and $h_{ijt}$ denote the corresponding
parameter. We seek to impose both temporal and spatial smoothness 
constraints on the parameters. Specifically, we seek a solution
for $h$ which is piecewise linear in time and piecewise constant in
space (although higher-order smoothness can be imposed with minimal
alterations to the methodology). We achieve this goal
by solving the following optimization problem: 
{\small
\begin{equation}
\label{eq:l1tf_var_st}.
\begin{aligned}
&\min_h \sum_{i,j,t}h_{ijt}+y_{ijt}^2e^{-h_{ijt}}\\& +\lambda_t
\sum_{i,j} \sum_{t=2}^{T-1}
\left|h_{ij(t-1)}-2h_{ijt}+h_{ij(t+1)}\right|\\ 
&+\lambda_s \sum_{t,j} \sum_{i=1}^{n_r-1} \left|h_{ijt}-h_{(i+1)jt}\right|
+\lambda_s \sum_{t,i} \sum_{j=1}^{n_c-1} \left|h_{ijt}-h_{i(j+1)t}\right|
\end{aligned}
\end{equation}
}%

@@

\begin{comment}
General (unconstrained) optimization problems $\min_z
\sum_i f_i(z)$, $z\in\mathbb{R}^n$ can be rewritten equivalently using a
collection of local variables, $x_i\in \mathbb{R}^{n_i}$, and a
constraint. Following the notation of 
\cite{boyd_distributed_2011}, let $k=\mathscr{G}(i,j)$ which
means that the $j^\text{th}$ entry of $x_i$ is $z_k$ (or
$(x_i)_j=z_k$) and define $\tilde{z}_i \in \mathbb{R}^{n_i}$ by
$(\tilde{z}_i)_j=(x_i)_j$. Then the original unconstrained
optimization problem is equivalent to the following constrained
problem
$\min_{\{x_1,...,x_N  \}} \sum_i f_i(x_i)$ subject to $\tilde{z}_i=x_i$.
Now, we can apply ADMM to the augmented Lagrangian of this
problem: 
\begin{align}
x_i&\leftarrow \argmin_{x_i} f_i(x_i) + (u_i)^\top x_i +
     (\rho/2) \norm{ x_i-\tilde{z}_i }_2^2\\
%\label{eq:ADMM_steps}
z_k&\leftarrow (1/S_k)\sum_{\mathscr{G}(i,j)=k} (x_i)_j\\
u_i&\leftarrow u_i + \rho (x_i-\tilde{z}_i)
\end{align}
\end{comment}

@@

Assume it is possible to define a set of local variables
$x_i \in \mathbb{R}^{n_i}$ such that $f(h)=\sum_i f_i(x_i)$, where
each $x_i$ is a subset of the global variable $h$. More specifically,
each entry of the local variables corresponds to an entry of the
global variable. Therefore we can define a mapping $\mathscr{G}(i,j)$
from the local variable indices into the global variable indices:
$k=\mathscr{G}(i,j)$ means that the $j^\text{th}$ entry of $x_i$ is
$h_k$ (or $(x_i)_j=h_k$). For ease of notation, define $\tilde{h}_i
\in \mathbb{R}^{n_i}$ as $(\tilde{h}_i)_j=h_{\mathscr{G}(i,j)}$. Then,
the original optimization problem is equivalent to:  
\begin{equation}
\begin{aligned}
&\min_{\{x_1,...,x_N  \}} \sum_{i=1}^N f_i(x_i) &
s.t. &\quad \tilde{h}_i=x_i.
\end{aligned}
\label{eq:consADMM}
\end{equation}
It is important to note that each entry of the global variable may
correspond to several entries of the local variables and so the
constraints $\tilde{h}_i=x_i$ enforce consensus between the local
variables corresponding to the same global variable.  
The augmented Lagrangian corresponding to 
\eqref{eq:consADMM} is $L_\rho(x,h,y)=\sum_i
\big(f_i(x_i)+u_i^\top(x_i-\tilde{h}_i) + (\rho/2) \lVert
x_i-\tilde{h}_i \lVert_2^2 \big)$. Now, we can apply ADMM to
$L_\rho$. This results in solving $N$ independent optimization
problems followed by a step to achieve consensus among the solutions
in each iteration. 
% which results in the following ADMM updates:
% \begin{align}
% x_i&\leftarrow\argmin_{x_i} f_i(x_i) + (u_i)^\top x_i +
%      (\rho/2)  \norm{ x_i-\tilde{h}_i }_2^2\\
% h_k&\leftarrow (1/S_k)\sum_{\mathscr{G}(i,j)=k} (x_i)_j \label{eq:ADMM_steps}\\
% u_i&\leftarrow u_i + \rho (x_i-\tilde{h}_i).
% \end{align}
% \attn{If you hate the left arrow, that's fine. But change it
%   everywhere, and use $=$. The $:=$ notation means that ``thing on
%   left is defined to be thing on right'' and isn't really appropriate here.}
% Here, $S_k$ is the number of local variable entries that correspond to
% $h_k$, and $u_i$ are the Lagrange multipliers.  
To solve the optimization problem \eqref{eq:l1tf_var_st_mat} using
this method, we need to address two 
questions: first, how to choose the 
local variables $x_i$, and second, how to the update them.
\begin{figure}[tb]
  \centering
  \includegraphics[width=.6\columnwidth]{Figures/data_cube}
  \caption{The cube represents the global variable $h$ in space and
    time. The sub-cubes specified by the white lines are
    $x_i$.}
  \label{fig:data_cube}
\end{figure}

@@

\begin{algorithm}[tb]
  \caption{Consensus ADMM }
  \label{alg:conADMM}
  \begin{algorithmic}[1]
    \STATE {\bfseries Input:} data $y$, penalty matrix $D$, 
    $\epsilon, \rho,\lambda_t,\lambda_s >0$.
    \STATE {\bf Set:} $h\leftarrow 0$, $z\leftarrow 0$, $u\leftarrow
    0$. \COMMENT{Initialization} 
    \REPEAT
    \STATE $\begin{aligned}x_i&\leftarrow\argmin_{x_i} -l(y_i\given
    x_i)+\Lambda_{(i)}^\top |D_{(i)}x_i|\\&\quad+ (u_i)^\top x_i +
    (\rho/2)  \lVert x_i-\tilde{h}_i \rVert_2^2\end{aligned}$. \\\COMMENT{Update local
      vars using PDIP}
    \STATE $h_k\leftarrow (1/S_k)\sum_{\mathscr{G}(i,j)=k} (x_i)_j
    $. \COMMENT{Global update.}
    \STATE $ u_i\leftarrow u_i + \rho (x_i-\tilde{h}_i)$. \COMMENT{Dual update}
    \UNTIL {$\max\left\{\norm{h^{m+1}-h^m},\ \norm{h^m-x^m}\right\} < \epsilon$}
    \STATE {\bf Return:} $h$.
  \end{algorithmic}
\end{algorithm}

@@

Consider the generic optimization problem
$\min_x f(x)+g(Dx)$
where $x\in \mathbb{R}^n$ and $D\in \mathbb{R}^{m\times n}$. Each
iteration of the linearized ADMM
algorithm~\cite{parikh_proximal_2014} for solving this problem 
has the form
\begin{equation}
  \label{eq:linADMM_steps}
  \begin{aligned}
    x & \leftarrow \prox_{\mu f} \left(x - (\mu/\rho)D^\top (D x - z + u )\right)\\
    z & \leftarrow \prox_{\rho g} \left(z + u\right)\\
    u & \leftarrow u + D x - z
  \end{aligned}
\end{equation}
where the algorithm parameters $\mu$ and $\rho$ satisfy $0 < \mu <
\rho/\norm{D}_2^2$, $z,u\in \mathbb{R}^m$ and the proximal operator is
defined as $\prox_{\alpha \varphi}(u) = \min_x \,\, \alpha \cdot
\varphi(x)+\frac{1}{2} \norm{ x-u}_2^2$. Proximal algorithms are feasible
when these proximal operators can be evaluated efficiently which, as
we show next, is the case.

@@

\begin{lemma}
  Let $f(x) = \sum_{k} x_{k} + y_{k}^2e^{-x_{k}}$ and $g(x) =
  \norm{x}_1$. Then, 
  \begin{equation}
    \begin{aligned}
      [\prox_{\mu f}(u)]_k &= \mathscr{W}\bigg(\frac{y_k^2}{\mu}
      \exp\bigg(\frac{1-\mu u_k}{\mu}\bigg) \bigg) + \frac{1-\mu u_k}{\mu},\\
      \prox_{\rho g}(u) &= S_{\rho\lambda}(u)
    \end{aligned}
  \end{equation}
  where $\mathscr{W}(\cdot)$ is the Lambert W function
  \cite{corless_lambertw_1996},  $[S_{\alpha}(u)]_k = \sign(u_k)(|u_k|
  -\alpha_k)_+$ and $(v)_+=v\vee 0$.
\end{lemma}
% \begin{proof}
%   If $f(x)=\sum_k f_k(x_k)$ then $[\prox_{\mu f}(x)]_k =
%   \prox_{\mu f_k}(u_k)$. So 
%   $[\prox_{\mu f}(u)]_k=\min_{x_k} \,\,
%   \mu\big(x_k+y_{k}^2e^{-x_{k}}\big)+\frac{1}{2}  (x_k-u_k)^2.$
%   Setting the derivative to 0 and solving for $u_k$ gives the
%   result. Similarly, $[\prox_{\rho g}(u)]_\ell=\rho
%   \lambda_\ell |z_\ell|+1/2(z_\ell-u_\ell)^2$. This is not differentiable,
%   but the solution must satisfy $\rho \cdot \lambda_\ell \cdot \partial
%   \big(|z_\ell| \big)=u_\ell-z_\ell$ where $\partial \big(|z_\ell| \big)$ is the
%   sub-differential of $|z_\ell|$. The solution is the soft-thresholding
%   operator $S_{\rho\lambda_\ell}(u_\ell)$.
% \end{proof}
The proof is fairly straightforward and given
in~\autoref{sec:proof-lemma-1} in the Supplement.  
Therefore, \autoref{alg:linADMM} gives a different method for solving
the same problem. In this case, both the primal update and the
soft thresholding step are performed elementwise at each point of the
spatio-temporal grid. It can therefore be extremely fast to perform
these steps. However, because there are now many more dual variables,
this algorithm will require more outer iterations to achieve
consensus. It therefore is highly problem and architecture dependent
whether \autoref{alg:conADMM} or \autoref{alg:linADMM} will be more
useful in any particular context. In our experience,
\autoref{alg:conADMM} requires an order of magnitude fewer iterations, but
each iteration is much slower unless carefully parallelized.

@@

Before examining real data, we apply our model to some synthetic
data. This example was constructed to mimic the types of spatial and
temporal phenomena observable in typical climate data. We generate a
complete spatio-temporal field wherein 
observations at all time steps and all locations are
independent Gaussian random variables with zero mean. However, the
variance of these random variables follows a smoothly varying function
in time and space given by the following parametric model:
\begin{align}
  \sigma^2(t,r,c) & =\sum_{k=1}^{K} W_k(t) \cdot \exp\bigg(
                    \frac{(r-r_k)^2+(c-c_k)^2} {2\sigma_k^2} \bigg)\\
  W_k(t) & =\alpha_k \cdot t + \exp(\sin(2\pi\omega_k t+\phi_k)) .
\label{eq:sourceVar}
\end{align}
The variance at each time and location is computed as the
weighted sum of $K$ bell-shaped functions where the weights are
time-varying, consist of a linear trend %$\alpha_k \cdot t$ 
and a periodic term. 
%$\beta_k \cdot \sin(2\pi\omega_k t+\phi_k)$. 
The
bell-shaped functions impose spatial smoothness while the linear
trend and the periodic terms enforce the temporal smoothness similar
to the seasonal component in real climate data. We simulated the
data on a 5$\times$7 grid for 780 time steps with $K=4$. This yields a
small enough problem to be evaluated many times while
still mimicking important properties of climate data.
Specific parameter choices of the variance function are shown in
\autoref{tab:sim_params}.
For illustration, we also plot the
variance function for all locations at $t=25$ and $t=45$
(\autoref{fig:true_var_spatial}, top panel) as well as
the variance across time at location $(0,0)$ (\autoref{fig:true_var_spatial}, bottom panel, orange).
\begin{table}[tb]
  \caption{Parameters used to simulate data.}
  \label{tab:sim_params}
  \begin{center}
    \begin{tabular}{ccccccc}
      \hline
      $s$ & $r_s$ & $c_s$ & $\sigma_s$ &$\alpha_s$ & $\omega_s$ & $\phi_s$\\
      \hline
      1 & 0 & 0 & 5 & 0.5 & 0.121 & 0 \\
      2 & 0 & 5 & 5 & 0.1 & 0.121 & 0 \\
      3 & 3 & 0 & 5 & -0.5 & 0.121 & $\pi/2$ \\
      4 & 3 & 5 & 5 & -0.1 & 0.121 & $\pi/2$ \\
      \hline
    \end{tabular}
  \end{center}
\end{table} 
\begin{figure}[tb]
  \centering	
  \includegraphics[height=.1\textheight]{Figures/true_var_spatial}\\
  \includegraphics[height=.15\textheight]{Figures/true_fitted_var}
  \caption{Top: Variance function at $t=25$ (left) and $t=45$
    (right). Bottom: The true (orange) and estimated standard deviation 
    function at the location (0,0). The estimated values are
    obtained using linearized ADMM with $\lambda_s=0.1$ and two
    values of $\lambda_t$: $\lambda_t=5$ (blue) and
    $\lambda_t=100$ (green).} \label{fig:true_var_spatial}
\end{figure}

>>>>

%Arash
\usepackage{amsmath,amsthm}
\usepackage{amssymb}
\usepackage{mathrsfs}
\DeclareMathOperator*{\argmin}{argmin}
%\usepackage{siunitx}
%\usepackage{comment}

@@

%\usepackage{autonum}

@@

Trends in terrestrial temperature variability are perhaps more relevant for species
viability than trends in mean
temperature~\cite{huntingford_no_2013},
because an increase in
temperature variability will increase the probability of extreme hot
or cold
outliers~\cite{VasseurDeLong2014}. Recent climate literature
suggests that it is more difficult for society to adapt to these
extremes than to the gradual increase in the mean temperature
\cite{hansen_perception_2012,huntingford_no_2013}. Furthermore, the willingness of popular media to
emphasize the prevalence extreme cold events coupled with a
fundamental misunderstanding of the relationship between climate (the global
distribution of weather over the long run) and weather (observed
short-term, localized behavior) leads to public misunderstanding of climate
change. In fact, a point of active debate is the extent to which the observed
increased frequency of extreme cold events in the 
northern hemisphere can be attributed to increases in temperature
variance  rather than to changes in
mean climate~\cite{Screen2014,FischerBeyerle2013,TrenberthZhang2014}.

Nevertheless, research examining trends in the volatility of
spatio-temporal climate data is scarce. \citet{hansen_perception_2012} studied the change in
the standard deviation (SD) of the surface temperature in the NASA
Goddard Institute for Space Studies gridded temperature dataset by
examining the empirical SD at each spatial location relative to that
location's SD over a base period and showed
that these estimates are increasing.
\citet{huntingford_no_2013} took a similar
approach in analyzing the ERA-40 data set. They argued that, while there
is an increase in the SDs from 1958-1970 to 1991-2001, it
is much smaller than found by
\citet{hansen_perception_2012}. \citet{huntingford_no_2013} also computed the
time-evolving global SD from the detrended time-series at each
position and argued that the global SD has been
stable.

@@

While the motivation for our methodology is its application to large, gridded climate
data, we note that our algorithms are easily generalizable to
spatio-temporal data under convex loss, e.g. exponential family
likelihood. Furthermore the spatial structure can be broadly construed
to include general graph dependencies. Our current application uses
Gamma likelihood which lends itself well to modeling trends in
pollutant emissions or in astronomical phenomena like microwave
background radiation. Volatility estimation in oil and natural gas markets
or with financial data is another possibility. Our methods can also
be applied to resting-state fMRI data (though the penalty structure changes).

@@

Inspired by the $\ell_1$-trend filtering algorithm, we propose a
non-parametric model for estimating the variance of a time-series. To
this end, we assume that at each time step $t$, there is a parameter
$h_t$ such that the observations $y_t$
are independent normal variables with zero mean and variance
$\exp(h_t)$. The negative log-likelihood of the observed data in this
model is $l(y\given h) \propto -\sum_t h_t - y_t^2e^{-h_t}$. Crucially,
we assume that the parameters $h_t$ vary smoothly and estimate them by
minimizing the penalized, negative
log-likelihood:  
\begin{equation}
  \label{eq:l1tf_var}
  \min_h -l(y\given h)+\lambda \norm{ D_th }_1
\end{equation}
where $D_t$ has the same structure as above.

@@

@@

At a specific time $t$, the data are measured on a grid of points with
$n_r$ rows and $n_c$ columns for a total of $S=n_r\times n_c$ spatial
locations. Let $y_{ijt}$ denote the value of the 
observation at time $t$ on the $i^\text{th}$ row and $j^\text{th}$
column of the grid, and $h_{ijt}$ denote the corresponding
parameter. We seek to impose both temporal and spatial smoothness 
constraints on the parameters. Specifically, we seek a solution
for $h$ which is piecewise linear in time and piecewise constant in
space (although higher-order smoothness can be imposed with minimal
alterations to the methodology). We achieve this goal
by solving the following optimization problem: 
{\small
\begin{align}
  &\min_h \sum_{i,j,t}h_{ijt}+y_{ijt}^2e^{-h_{ijt}}\notag\\
  & +\lambda_t \sum_{i,j} \sum_{t=2}^{T-1}
    \left|h_{ij(t-1)}-2h_{ijt}+h_{ij(t+1)}\right|\label{eq:l1tf_var_st}\\ 
  &+\lambda_s \sum_{t,j} \sum_{i=1}^{n_r-1} \left|h_{ijt}-h_{(i+1)jt}\right|
    +\lambda_s \sum_{t,i} \sum_{j=1}^{n_c-1}
    \left|h_{ijt}-h_{i(j+1)t}\right|\notag
\end{align}
}%

@@

@@

Assume it is possible to define a set of local variables
$x_i \in \mathbb{R}^{n_i}$ such that $f(h)=\sum_i f_i(x_i)$, where
each $x_i$ is a subset of the global variable $h$. More specifically,
each entry of the local variables corresponds to an entry of the
global variable. Therefore we can define a mapping $G(i,j)$
from the local variable indices into the global variable indices:
$k=G(i,j)$ means that the $j^\text{th}$ entry of $x_i$ is
$h_k$ (or $(x_i)_j=h_k$). For ease of notation, define $\tilde{h}_i
\in \mathbb{R}^{n_i}$ as $(\tilde{h}_i)_j=h_{G(i,j)}$. Then,
the original optimization problem is equivalent to:  
\begin{equation}
\begin{aligned}
&\min_{\{x_1,...,x_N  \}} \sum_{i=1}^N f_i(x_i) &
s.t. &\quad \tilde{h}_i=x_i.
\end{aligned}
\label{eq:consADMM}
\end{equation}
It is important to note that each entry of the global variable may
correspond to several entries of the local variables and so the
constraints $\tilde{h}_i=x_i$ enforce consensus between the local
variables corresponding to the same global variable.  
The augmented Lagrangian corresponding to 
\eqref{eq:consADMM} is $L_\rho(x,h,y)=\sum_i
\big(f_i(x_i)+u_i^\top(x_i-\tilde{h}_i) + (\rho/2) \lVert
x_i-\tilde{h}_i \lVert_2^2 \big)$. Now, we can apply ADMM to
$L_\rho$. This results in solving $N$ independent optimization
problems followed by a step to achieve consensus among the solutions
in each iteration. 
% which results in the following ADMM updates:
% \begin{align}
% x_i&\leftarrow\argmin_{x_i} f_i(x_i) + (u_i)^\top x_i +
%      (\rho/2)  \norm{ x_i-\tilde{h}_i }_2^2\\
% h_k&\leftarrow (1/S_k)\sum_{\mathscr{G}(i,j)=k} (x_i)_j \label{eq:ADMM_steps}\\
% u_i&\leftarrow u_i + \rho (x_i-\tilde{h}_i).
% \end{align}
% \attn{If you hate the left arrow, that's fine. But change it
%   everywhere, and use $=$. The $:=$ notation means that ``thing on
%   left is defined to be thing on right'' and isn't really appropriate here.}
% Here, $S_k$ is the number of local variable entries that correspond to
% $h_k$, and $u_i$ are the Lagrange multipliers.  
To solve the optimization problem \eqref{eq:l1tf_var_st_mat} using
this method, we need to address two 
questions: first, how to choose the 
local variables $x_i$, and second, how to the update them.
\begin{figure}[tb]
  \centering
  \includegraphics[width=.6\columnwidth]{Figures/data_cube}
  \caption{The cube represents the global variable $h$ in space and
    time. The sub-cubes specified by the white lines are
    $x_i$.}
  \label{fig:data_cube}
\end{figure}

@@

\begin{algorithm}[tb]
  \caption{Consensus ADMM }
  \label{alg:conADMM}
  \begin{algorithmic}[1]
    \STATE {\bfseries Input:} data $y$, penalty matrix $D$, 
    $\epsilon, \rho,\lambda_t,\lambda_s >0$.
    \STATE {\bf Set:} $h\leftarrow 0$, $z\leftarrow 0$, $u\leftarrow
    0$. \COMMENT{Initialization} 
    \REPEAT
    \STATE $\begin{aligned}x_i&\leftarrow\argmin_{x_i} -l(y_i\given
    x_i)+\Lambda_{(i)}^\top |D_{(i)}x_i|\\&\quad+ (u_i)^\top x_i +
    (\rho/2)  \lVert x_i-\tilde{h}_i \rVert_2^2\end{aligned}$. \\\COMMENT{Update local
      vars using PDIP}
    \STATE $h_k\leftarrow (1/S_k)\sum_{G(i,j)=k} (x_i)_j
    $. \COMMENT{Global update.}
    \STATE $ u_i\leftarrow u_i + \rho (x_i-\tilde{h}_i)$. \COMMENT{Dual update}
    \UNTIL {$\max\left\{\norm{h^{m+1}-h^m},\ \norm{h^m-x^m}\right\} < \epsilon$}
    \STATE {\bf Return:} $h$.
  \end{algorithmic}
\end{algorithm}

@@

Consider the generic optimization problem
$\min_x f(x)+g(Dx)$
where $x\in \mathbb{R}^n$ and $D\in \mathbb{R}^{m\times n}$. Each
iteration of the linearized ADMM
algorithm~\cite{parikh_proximal_2014} for solving this problem 
has the form
\begin{equation*}
  \label{eq:linADMM_steps}
  \begin{aligned}
    x & \leftarrow \prox_{\mu f} \left(x - (\mu/\rho)D^\top (D x - z + u )\right)\\
    z & \leftarrow \prox_{\rho g} \left(z + u\right)\\
    u & \leftarrow u + D x - z
  \end{aligned}
\end{equation*}
where the algorithm parameters $\mu$ and $\rho$ satisfy $0 < \mu <
\rho/\norm{D}_2^2$, $z,u\in \mathbb{R}^m$ and the proximal operator is
defined as $\prox_{\alpha \varphi}(u) = \min_x \,\, \alpha \cdot
\varphi(x)+\frac{1}{2} \norm{ x-u}_2^2$. Proximal algorithms are feasible
when these proximal operators can be evaluated efficiently which, as
we show next, is the case.

@@

\begin{lemma}
  Let $f(x) = \sum_{k} x_{k} + y_{k}^2e^{-x_{k}}$ and $g(x) =
  \norm{x}_1$. Then, 
  \begin{equation*}
    \begin{aligned}
      \bigl[\prox_{\mu f}(u) \bigl]_k &= \mathscr{W}\bigg(\frac{y_k^2}{\mu}
      \exp\bigg(\frac{1-\mu u_k}{\mu}\bigg) \bigg) + \frac{1-\mu u_k}{\mu},\\
      \prox_{\rho g}(u) &= S_{\rho\lambda}(u)
    \end{aligned}
  \end{equation*}
  where $\mathscr{W}(\cdot)$ is the Lambert W function
  \cite{corless_lambertw_1996},  $[S_{\alpha}(u)]_k = \sign(u_k)(|u_k|
  -\alpha_k)_+$ and $(v)_+=v\vee 0$.
\end{lemma}
% \begin{proof}
%   If $f(x)=\sum_k f_k(x_k)$ then $[\prox_{\mu f}(x)]_k =
%   \prox_{\mu f_k}(u_k)$. So 
%   $[\prox_{\mu f}(u)]_k=\min_{x_k} \,\,
%   \mu\big(x_k+y_{k}^2e^{-x_{k}}\big)+\frac{1}{2}  (x_k-u_k)^2.$
%   Setting the derivative to 0 and solving for $u_k$ gives the
%   result. Similarly, $[\prox_{\rho g}(u)]_\ell=\rho
%   \lambda_\ell |z_\ell|+1/2(z_\ell-u_\ell)^2$. This is not differentiable,
%   but the solution must satisfy $\rho \cdot \lambda_\ell \cdot \partial
%   \big(|z_\ell| \big)=u_\ell-z_\ell$ where $\partial \big(|z_\ell| \big)$ is the
%   sub-differential of $|z_\ell|$. The solution is the soft-thresholding
%   operator $S_{\rho\lambda_\ell}(u_\ell)$.
% \end{proof}
The proof is fairly straightforward and given
in~\autoref{sec:proof-lemma-1} in the Supplement.  
Therefore, \autoref{alg:linADMM} gives a different method for solving
the same problem. In this case, both the primal update and the
soft thresholding step are performed elementwise at each point of the
spatio-temporal grid. It can therefore be extremely fast to perform
these steps. However, because there are now many more dual variables,
this algorithm will require more outer iterations to achieve
consensus. It therefore is highly problem and architecture dependent
whether \autoref{alg:conADMM} or \autoref{alg:linADMM} will be more
useful in any particular context. In our experience,
\autoref{alg:conADMM} requires an order of magnitude fewer iterations, but
each iteration is much slower unless carefully parallelized.

@@

Before examining real data, we apply our model to some synthetic
data. This example was constructed to mimic the types of spatial and
temporal phenomena observable in typical climate data. We generate a
complete spatio-temporal field wherein 
observations at all time steps and all locations are
independent Gaussian random variables with zero mean. However, the
variance of these random variables follows a smoothly varying function
in time and space given by the following parametric model:
\begin{align*}
  \sigma^2(t,r,c) & =\sum_{k=1}^{K} W_k(t) \cdot \exp\bigg(
                    \frac{(r-r_k)^2+(c-c_k)^2} {2\sigma_k^2} \bigg)\\
  W_k(t) & =\alpha_k \cdot t + \exp(\sin(2\pi\omega_k t+\phi_k)) .
\label{eq:sourceVar}
\end{align*}
The variance at each time and location is computed as the
weighted sum of $K$ bell-shaped functions where the weights are
time-varying, consist of a linear trend %$\alpha_k \cdot t$ 
and a periodic term. 
%$\beta_k \cdot \sin(2\pi\omega_k t+\phi_k)$. 
The
bell-shaped functions impose spatial smoothness while the linear
trend and the periodic terms enforce the temporal smoothness similar
to the seasonal component in real climate data. We simulated the
data on a 5$\times$7 grid for 780 time steps with $K=4$. This yields a
small enough problem to be evaluated many times while
still mimicking important properties of climate data.
Specific parameter choices of the variance function are shown in
\autoref{tab:sim_params}.
For illustration, we also plot the
variance function for all locations at $t=25$ and $t=45$
(\autoref{fig:true_var_spatial}, top panel) as well as
the variance across time at location $(0,0)$ (\autoref{fig:true_var_spatial}, bottom panel, orange).
\begin{table}[tb]
  \caption{Parameters used to simulate data.}
  \label{tab:sim_params}
  \begin{center}
    \begin{tabular}{ccccccc}
      \hline
      $s$ & $r_s$ & $c_s$ & $\sigma_s$ &$\alpha_s$ & $\omega_s$ & $\phi_s$\\
      \hline
      1 & 0 & 0 & 5 & 0.5 & 0.121 & 0 \\
      2 & 0 & 5 & 5 & 0.1 & 0.121 & 0 \\
      3 & 3 & 0 & 5 & -0.5 & 0.121 & $\pi/2$ \\
      4 & 3 & 5 & 5 & -0.1 & 0.121 & $\pi/2$ \\
      \hline
    \end{tabular}
  \end{center}
\end{table} 
\begin{figure}[tb]
  \centering	
  \includegraphics[height=.1\textheight]{Figures/true_var_spatial}\\
  \includegraphics[height=.15\textheight]{Figures/true_fitted_var}
  \caption{Top: Variance function at $t=25$ (left) and $t=45$
    (right). Bottom: The true (orange) and estimated standard deviation 
    function at the location (0,0). The estimated values are
    obtained using linearized ADMM with $\lambda_s=0.1$ and two
    values of $\lambda_t$: $\lambda_t=5$ (blue) and
    $\lambda_t=100$ (green).} \label{fig:true_var_spatial}
\end{figure}

We estimated the linearized ADMM for all combinations of values of
$\lambda_t$ and $\lambda_s$ from the sets $\lambda_t \in
\{0,1,5,10,50,100\}$ and $\lambda_s \in \{0,0.05,0.1,0.2,0.3\}$. For
each pair, we then compute the mean absolute error (MAE) between the
estimated variance and the true variance at all locations and all time
steps. For $\lambda_t=5$ and $\lambda_s=0.1$, the MAE was minimized. 
\begin{figure}[tb]
  \centering	
  \includegraphics[width=.48\columnwidth]{Figures/convergence}  
  \includegraphics[width=.48\columnwidth]{Figures/modelComp_MAE}  
  \caption{Left: Value of the objective function for linearized (orange)
    and consensus (blue) ADMM against iteration. Right: MAE for (1) our
    method with optimal values of $\lambda_t$ and $\lambda_s$ (2)
    spatial penalty only (3) temporal
    penalty only and (4) a GARCH(1,1).} 
  \label{fig:true_fitted_var}
\end{figure}
The
bottom panel of \autoref{fig:true_var_spatial} shows the true and the
estimated standard deviation at location (0,0) 
and $\lambda_t=5$ (blue) and $\lambda_t=100$ (green) ($\lambda_s=0.1$).
Larger values of $\lambda_t$ lead to estimated values
which are ``too smooth". The left panel of \autoref{fig:true_fitted_var} shows the
convergence of both Algorithms as a function of iteration. It is
important to note that each iteration of the linearized
algorithm takes 0.01 seconds on average while each iteration of the
consensus ADMM takes about 20 seconds. Thus, where the lines meet at
400 iterations requires about 4 seconds for the linearized method and
2 hours for the consensus method. For consensus ADMM, computation per
iteration per core requires $\sim$10 seconds with $\sim$4 seconds
for communication. In general, the literature suggests linear
convergence for ADMM~\citep{NishiharaLessard2015} and~\autoref{fig:true_fitted_var} seems to fit in the linear framework for both algorithms, though with different constants.
